---
title: "Chipotle Meals are Nutrient Dense"
author: "[Cody Stumpo](https://github.com/CodyStumpo/)"
date: "February 28, 2015"
output: html_document
---
## Full Meals

The NY Times got ahold of some detailed Chipotle order data and '[analyzed](http://www.nytimes.com/interactive/2015/02/17/upshot/what-do-people-actually-order-at-chipotle.html?abt=0002&abg=1&_r=0)' it.  They concluded the meal has a lot of calories, sodium, and saturated fat.
Calories you need, obviously.  If you need about 2000 of them a day and once in a while you eat a big meal, that meal might be 1000 or even 1200 calories.  Unless you eat a real breakfast, Mega-Chipotle for lunch and Mega-Chipotle again for dinner, and then go back and do it all again the next day, you're not creating any caloric problems for yourself.  So stop feeling bad about that.  The question is really whether you're making good use of the calories you're getting.  Two things scientists used to think were possibly bad for you were sodium and saturated fat.  But they don't necessarily agree on that any more.  The only things well and truly bad for you by unanimous consent are trans fats and sugar.  Chipotle has almost none of these in any of its foods.  Ironically, the least healthy meal there now is the salad, since the dressing has a little sugar (not much).

I didn't add chips or sodas to meals, just analyzed meals (Burritos, Tacos, Bowls, Salads). The data is great because it has the full details of every component in the meal (e.g. a steak burrito with white rice, pinto beans, hot salsa, and guacamole). And Chipotle gives us [quite a bit of info](http://www.chipotle.com/en-US/menu/nutrition_calculator/nutrition_calculator.aspx) about each component. Let's take a look first at meals through the food-shaming lens of the NY Times. 


```{r, echo=FALSE}
setwd("/Users/cstumpo/Box Sync/cms/RProjects/Chipotle_Nutrition")
load("out/full_meals.rda")
hist(full_meals$Calories)
hist(full_meals$Sugar.g.)
```

Chipotle sells nothing with any Trans Fat at all, so there's no chart to show.  Guidelines on sugar say to keep it under 90g a day, so even if you get the salad dressing or (very slightly) sugary salsa, you're not contributing to any significant accumulation of sugar.

So what are you getting for your 850 +/- 200 calories?  Well, first off let's analyze the percent of calories that come from fat vs. protein vs. carbs.  There's no universal right mix here.  It kind of depends on your goals. In the plots below, I also color the points to reflect how many calories the meal has overal.  Bright Green is lowest calories.  Bright Magenta is highest calories. It's a continuous scale between those two extremes.

```{r, echo = FALSE}
gmPal <- colorRampPalette(c('green','magenta')) #green = lo-cal, magenta = hi-cal
full_meals$CalColors <- gmPal(50)[as.numeric(cut(full_meals$Calories,breaks = 100))]

#par(pty="s") #make plots square
rng=c(0,0.8)

pairs(~fatByCal+proteinByCal+carbByCal, data = full_meals, col=full_meals$CalColors, 
      cex= 0.1, xlim=rng, ylim=rng, main="Chipotle Meals", sub="green = lo-cal; magenta=hi-cal")


```

More or less, you're going to get 30% of the calories from protein and then can trade off fat for protein in the rest.  You'll see a little cluster of 15% protein calorie meals. That's some vegetarian meals (the low-calorie ones) and some with lots of sides (the hi-calorie ones).

What about all the explicitly good stuff in the meal?  How nutrient dense are typical chipotle meals?  Suppose you really do aim to eat 2000 calories and 800 are coming at Chipotle (~40%).  Are you going to get 40% of your daily recommended value of Vitamin A, Vitmain C, Iron, Calcium, and Fiber?  Generally the answer is yes.  You are getting nutrient-dense food.

The diagonal line in each plot below represents a 1:1 ratio of % Daily Values of the nutrient to % Daily Value of Calories.  Each point is a meal.  A meal above the line is nutrient dense for that nutrient.  A meal far below the line has 'empty calories' as far as that nutrient is concerned.  Most people are putting together nutrient-dense meals.


```{r, echo=FALSE}
rng=c(0,1.5)
par(pty="s") #make plots square
palette(c("Orange","Grey","Brown", "Blue", "Green"))

plot(full_meals$"Calories Share", full_meals$"Protein.g. Share",
     xlim=rng, ylim=rng, xlab = "%DV Calories", ylab="%DV Protein", 
     main = "Chipotle Meal Protein / Calorie Density", col=full_meals$Main)
abline(a = 0, b=1)
legend('topright', legend = levels(full_meals$Main), col = 1:5, cex = 0.8, pch = 1)



plot(full_meals$"Calories Share", full_meals$"DietaryFiber.g. Share",
     xlim=rng, ylim=rng, xlab = "%DV Calories", ylab="%DV Fiber", 
     main = "Chipotle Meal Fiber / Calorie Density", col=full_meals$Main)
abline(a = 0, b=1)
legend('topright', legend = levels(full_meals$Main), col = 1:5, cex = 0.8, pch = 1)



plot(full_meals$"Calories Share", full_meals$"VitaminA..DV. Share",
     xlim=rng, ylim=rng, xlab = "%DV Calories", ylab="%DV Vitamin A", 
     main = "Chipotle Meal Vitamin A / Calorie Density", col=full_meals$Main)
abline(a = 0, b=1)
legend('topright', legend = levels(full_meals$Main), col = 1:5, cex = 0.8, pch = 1)



plot(full_meals$"Calories Share", full_meals$"VitaminC..DV. Share",
     xlim=rng, ylim=rng, xlab = "%DV Calories", ylab="%DV Vitamin C", 
     main = "Chipotle Meal Vitamin C / Calorie Density", col=full_meals$Main)
abline(a = 0, b=1)
legend('topright', legend = levels(full_meals$Main), col = 1:5, cex = 0.8, pch = 1)



plot(full_meals$"Calories Share", full_meals$"Iron..DV. Share",
     xlim=rng, ylim=rng, xlab = "%DV Calories", ylab="%DV Iron", 
     main = "Chipotle Meal Iron / Calorie Density", col=full_meals$Main)
abline(a = 0, b=1)
legend('topright', legend = levels(full_meals$Main), col = 1:5, cex = 0.8, pch = 1)


plot(full_meals$"Calories Share", full_meals$"Calcium...DV. Share",
     xlim=rng, ylim=rng, xlab = "%DV Calories", ylab="%DV Calcium", 
     main = "Chipotle Meal Calcium / Calorie Density", col=full_meals$Main)
abline(a = 0, b=1)
legend('topright', legend = levels(full_meals$Main), col = 1:5, cex = 0.8, pch = 1)
```

Finally, you are getting your money's worth.  The meal costs $10 but it is half of everything you need for the day.

## Individual Meal Components
It's also interesting to look at how each individual meal component you can order contributes to nutrition.  On the 12 dimensions of nutrition information Chipotle provides (accepting trans fats as zero across the board, remember), how much bang for your calorie buck do you get in each?  That is to say, for component i, 
 %DV_NutritionDimension[i] /  %DV_Calories[i].
 
 So for example, take Pinto Beans. Each ring of the spiderweb away from the innermost is an additional 1x on denseness.  The zeroth line is 0x, first line is 1x, the second line is 2x,... in terms of % of DV provided compared to % of DV Calories provided.  The actual Vitamins & Minerals pretty much the more the better.  The rest, don't take it too harmfully if something has 2x density for sodium or fat or something 'bad.' The DV in those cases are intended to be more like limits than minimums, but recent science is not backing up the idea that they need to be limited. For ease of reading, I cut off the axes at 400%.  Beyond that, you can just think of how dense the subject of the plot is in that nutrition element as 'very.' 
 
```{r, echo=FALSE, warning=FALSE}
load("out/nut_rich.rda")
library(fmsb)

i=10 #Pinto Beans
dat=nut_rich[i,4:15]
radarchart(rbind(rep(4,12), rep(0,12), dat), seg=4, title = nut_rich$Name[i])
```

Pinto Beans, per calorie, are an excellent source of Iron, Vitamin C, Protein, and Fiber.  But more interesting maybe is to compare that against some like options.

### Comparisons of options in each class

#### Beans

```{r, echo=FALSE, warning=FALSE}
palette(c("Black", "Brown"))
beans=nut_rich[c(9,10),]
radarchart(rbind(rep(4,12), rep(0,12), beans[, 4:15]), seg=4, title="Beans")
legend('topright', legend = beans$Name[1:2], col = 1:2, cex = 0.8, pch = 1)
```

They're very similar.  Pinto Beans have some Vitamin C that the Black Beans don't.

#### Rices

We could do the same for rices and see they too are quite similar to each other, but the white rice has some extra Vitamin A and Iron while the Brown rice offers more Fiber.

```{r, echo=FALSE, warning=FALSE}
palette(c("Green", "Brown"))
rices=nut_rich[c(7,8),]
radarchart(rbind(rep(4,12), rep(0,12), rices[, 4:15]), seg=4, title="Rices")
legend('topright', legend = rices$Name[1:2], col = 1:2, cex = 0.8, pch = 1)
```

#### Meats

All the Meats are of course excellent sources of Protein, with the Beefs also providing good Iron density.  

```{r, echo=FALSE, warning=FALSE}
palette(c("Orange","Grey","Brown", "Blue"))
meats=nut_rich[c(4,5,2,1),]
radarchart(rbind(rep(4,12), rep(0,12), meats[, 4:15]), seg=4, title="Meats")
legend('topright', legend = meats$Name[1:4], col = 1:4, cex = 0.8, pch = 1)
```

#### Fats

The Fatty sides are where you can add a lot of fat and calories (and flavor!) to your meal, if you're looking to do that.  They halfway hold their own in terms of the other nutritional elements.

```{r, echo=FALSE, warning=FALSE}
palette(c("Beige","Yellow","Green", "Orange"))
fats=nut_rich[c(16,17,18,24),]
radarchart(rbind(rep(4,12), rep(0,12), fats[, 4:15]), seg=4, title="Fats")
legend('topright', legend = fats$Name[1:4], col = 1:4, cex = 0.8, pch = 1)
```

#### Tortillas

Tortillas make a nice mouth feel to the whole meal, but limited nutrition for the calories.  The flour is a bit more nutritious than the corn.

```{r, echo=FALSE, warning=FALSE}
palette(c("Black","Yellow","Green", "Orange", "Blue"))
tortillas=nut_rich[c(6,20,21,22,23),]
radarchart(rbind(rep(4,12), rep(0,12), tortillas[, 4:15]), seg=4, title="Tortillas")
legend('topright', legend = tortillas$Name[1:5], col = 1:5, cex = 0.8, pch = 1)
```

#### Veggies
Vegetables are good for you.  You may have heard this before. They break these graphs, they are so nutrient-dense.  A lettuce serving is 5 calories (0.25% of your DV), and it has 35% of your DV for Vitamin A. That's a density of 140x.

```{r, echo=FALSE, warning=FALSE}
palette(c("Orange","Red","Yellow", "Green", "Brown", "Purple"))
veg=nut_rich[c(11,12,13,14,15,19),]
radarchart(rbind(rep(4,12), rep(0,12), veg[, 4:15]), seg=4, title="Vegetables")
legend('topright', legend = veg$Name[1:6], col = 1:6, cex = 0.8, pch = 1)
```

## Example Meals
With the introdicution of Sofritas, you can order something like [a million different meals](http://www.businessinsider.com/how-many-combinations-can-you-order-at-chipotle-2013-7) at Chipotle. When I go to Chipotle, I usually gravitate to one of a half-dozen meals, that between them use most of the possible ingredients.  Let's take some and put them through this analysis.  


### Steak Soft Tacos
If I don't want a big meal, I might put together something like street tacos.  For me that's
"Steak", "Lettuce", "Guacamole", "Red Tomatillo Salsa", "Tomato Salsa", "Soft Corn Tortillas."  That has **680 Calories, split 39% from Fat, 39% from Carbs, 22% from Protein**.  It has OK nutrition except for a shortage of Calcium.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
load("out/nutrition.rda")
load("out/nutrition2.rda")


mySteakTacos = c("Steak", "Lettuce", "Guacamole", "Red Tomatillo Salsa", "Tomato Salsa", "Soft Corn Tortillas")


library(dplyr)
myTots2=function(meal){
  ings.nut=filter(nutrition2, Name %in% meal)
  colSums(ings.nut[c(3:5,7:16)])
}


x = myTots2(mySteakTacos)
radarchart (data.frame(rbind(rep(4,12), rep(0,12), x[2:13]/x[1])), seg=4, title="Steak Tacos")

```


### Chicken Salad
Another meal I like is a Chicken Salad:  "Chicken", "Lettuce", "Vinaigrette", "Black Beans", "Corn Salsa", "Cheese."

That has **755 Calories, split 47% from Fat, 29% from Carbs, 24% from Protein**.  It is actually less nutritious per calorie than the Tacos, as the salad dressing offers nothing (but flavor) and Chicken doesn't offer as much Iron.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

myChickenSalad = c("Chicken", "Lettuce", "Vinaigrette", "Black Beans", "Corn Salsa", "Cheese")
x = myTots2(myChickenSalad)
radarchart (data.frame(rbind(rep(4,12), rep(0,12), x[2:13]/x[1])), seg=4, title= "Chicken Salad")

```


### Sofritas Bowl
Sometimes, I am just tagging along with some group, not raelly in the mood for Chipotle. That's when I might put together a Sofritas Bowl: "Sofritas", "Brown Rice", "Pinto Beans", "Sour Cream", "Corn Salsa", "Green Tomatillo Salsa."
That has **675 Calories, split 38% from Fat, 48% from Carbs, and 13% from Protein**.  It's also across the board nutritious.  While offering less Protein than most Chipotle meals, it still carries it's weight in terms of getting to my daily value.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

mySofritasBowl = c("Sofritas", "Brown Rice", "Pinto Beans", "Sour Cream", "Corn Salsa", "Green Tomatillo Salsa")
x = myTots2(mySofritasBowl)
radarchart (data.frame(rbind(rep(4,12), rep(0,12), x[2:13]/x[1])), seg=4, title= "Sofritas Bowl")

```


### Carnitas Burrito
Finally, I do sometimes go in whole hog and just get a Carnitas Burrito:
"Carnitas", "Pinto Beans", "White Rice", "Guacamole", "Green Tomatillo Salsa","Soft Flour Tortilla"
 "Sofritas", "Brown Rice", "Pinto Beans", "Sour Cream", "Corn Salsa", "Green Tomatillo Salsa."
That has **1070 Calories, split 42% from Fat, 42% from Carbs, and 17% from Protein**.  This meal isn't quite worth it, from a nutrition per Calorie point of view, but I only get this once in a blue moon.


```{r, echo=FALSE, warning=FALSE, message=FALSE}

myCarnitasBurrito = c("Carnitas", "Pinto Beans", "White Rice", "Guacamole", "Green Tomatillo Salsa","Soft Flour Tortilla")
x = myTots2(myCarnitasBurrito)
radarchart (data.frame(rbind(rep(4,12), rep(0,12), x[2:13]/x[1])), seg=4, title= "Carnitas Burrito")

```
